Translation of B Implementations to the LLVM: Control and Modularity

### 2013-09-04 translation.tex
Updated the specification of the translation to reflect evolution in handling modular design (more precisely: import clauses in implementations).

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\text{\Return}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{imp}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.

@@

        \B{see} & : & seq \B{Name} \\
        \B{imp} & : & seq \B{Impo} \\

@@

        \B{op} & : & seq \B{Oper} \\

@@

@@

@@

        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \alt \B{Skip} \\

@@

\DD{Restore \B{Case} and \B{CaseBr}.}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def} 
      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}
  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
    and \nt{name} correspond respectively to integer types, integer literals,
    labels and names. Choices are separated by \alt and optionally delimited by
    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
    elements of the annotated entity.}
  \label{fig:llvm-grammar}
\end{figure}

@@

@@

  The parameter list of the functions thus produced contain one item for each
  input and output of the corresponding operation, and one item of type
  \llvm{\%M\$state\$*}, to pass the address of the structure encoding the
  state of the implementation and its modification by the callee. The return
  type is always \llvm{void}.
\end{enumerate}
This approach assumes that information on external B components used in \B{M} is
available during translation. To fulfill this hypothesis, the translation
produces an ``interface'' artifact containing the information that the
translation of an importation of \B{M} in a client machine
requires. Figure~\ref{fig:skel} summarizes the translation principles by
providing the structure of both the interface and the LLVM implementation thus
produced.

\begin{figure}
  \begin{center}
    (a) Structure of the interface artifact produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{(references to imported components)} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{declare void @M\$init\$(\%M\$state\$*)} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{declare void @M\$op(\%M\$state\$*, \ListOf{\nt{type}})} \\
    \end{tabbing}
    (b) Structure of the LLVM module produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer} \\
      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}}
    \end{tabbing}
  \end{center}
  \caption{Skeleton of the artifacts produced by the translation.}
  \label{fig:skel}
\end{figure}

@@

Code generation is defined with a set of rules defined by structural
recursion. For each grammatical construct, a code generation function is
defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
instructions, $\trad{~}{\B{Expr}}$ for expressions and so forth. Each such
function usually produces a text corresponding to bits of LLVM source code. It
may have additional parameter and results. The symbol $\sep$ is used to clearly
separate composed results. Auxiliary functions are introduced for
conveniency. To abbreviate the presentation, we denote
$\trad{~}{\ListOf{\B{A}}}$ as the translation function for sequences of \B{A}
elements, defined as the concatenation of the application of $\trad{~}{\B{A}}$
to the elements of the sequence.

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name. Also $\state(\B{M})
\isdef \llvm{"\%\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
state of component \B{M}. We assume an unlimited supply of labels and local
variables: we use $\lbl$ to return a fresh LLVM label and $\variable$ for a
fresh local variable name.

@@

The translation of a B module produces two artifacts: an interface and an (LLVM)
implementation. In the following, we describe how they are produced.

\subsection{Interface}

The definition of the interface $\intf(\B{M})$ of a component \B{M} has four
key elements, as shown in figure~\ref{fig:skel} (b):
\begin{enumerate}
\item $\intf(\B{M})_u$ is the list of used components: $\intf(\B{M})_u \isdef \B{M.imp}$.

\item $\intf(\B{M})_t$ is a textual description of the type representing the
  state of the component, i.e. a structure with one element per imported
  component and variable:
\begin{small}
$$\intf(\B{M})_t \isdef \llvm{"\{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
\end{small}
where $\trad{\B{i}}{\B{Impo}}_t \isdef \llvm{\state(\B{i})}$ and
$\trad{\B{v}}{\B{Vari}}_t \isdef \trad{\B{v.type}}{\B{Type}}$.

\item $\intf(\B{M})_o$ represents the operation declarations. It maps
operation names to the list of input parameter types and output parameter
types:
$\intf(\B{M})_o \isdef \trad{\B{M.op}}{\ListOf{\B{Oper}}}_t$ and
$\trad{\B{o}}{\B{Oper}}_t \isdef \B{o} \mapsto ( \trad{\B{o.inp}}{\ListOf{inp}} \sep \trad{\B{o.out}}{\ListOf{out}} )$, where
$\trad{\B{i}}{inp} \isdef \trad{\B{i.type}}{\B{Type}}$ and
$\trad{\B{o}}{out} \isdef \llvm{"\PH{\trad{\B{o.type}}{\B{Type}}}*"}$.
\end{enumerate}

\subsection{Implementation}

The actual LLVM source code is generated following the template shown in
figure~\ref{fig:skel} (a), as follows:
\begin{align*}
  \lefteqn{\trad{\B{i}, \intf}{\B{Impl}} \isdef
  \llvm{"\PH{\trad{\B{i.imp}, \intf}{\ListOf{IType}}}} 
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{IInit}}}
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{ICons}}}} \\
& \llvm{\PH{\trad{\B{i.imp}, \intf}{\ListOf{IOper}}}
  \quad \PH{\trad{\B{i.id}, \B{i.const}}{\ListOf{\B{Cons}}}}
  \quad \PH{\trad{\B{i}}{State}}} \\
& \llvm{@\PH{\B{i.id}}\$self\$ = common global \PH{\state(\B{i.id})} zeroinitializer \nl} \\
& \llvm{\PH{\trad{\B{i.init}}{Init}} 
  \quad \PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
\end{align*}
We shall now define each rule referenced in this template.
\begin{enumerate}
\item For the type definitions corresponding to the state of the imported
  components: $\trad{i, \intf}{IType} \isdef \llvm{"\PH{\state(\B{i})} =
    type \PH{\intf(\B{i})_t}\nl"}$.
\item For the declarations of the initialization functions in the imported
  components: $\trad{\B{i}, \intf}{Init} \isdef \llvm{"declare void
    @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
\item For the declarations of the imported operations: $\trad{\B{i},
    \intf}{IOper} = \trad{\B{i.id}, $\intf(\B{i})_o$}{\ListOf{Oper}}$ where
  $\trad{\B{Q}, $\B{op} \mapsto (i, o)$}{Oper} = \llvm{"declare void
    \PH{\Global(\B{op})}(\PH{\state(\B{Q})}*,\PH{i},\PH{o})"}$.
\item The translation of constant definitions is presented in section~\ref{sec:data}.
\item For the definition of the type for component states: \\
\noindent$\trad{\B{M}}{State} \isdef
\llvm{"\PH{\state(\B{M})} = type \PH{\intf(\B{M})_t}\nl"}$.
\item For the initialization, the definition of a function is generated. It
  first initializes the imported modules and then the current module (see
  section~\ref{sec:control}).
\item For each operation, a function definition is generated (see
  section~\ref{sec:control}).
\end{enumerate}
Let $\idx$ be a function that maps each imported machine and state variable to
its position (starting from zero).
\section{Translation: Data}
\label{sec:data}

>>>>

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\text{\Return}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{impo}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.

@@

        \B{see} & : & seq \B{Mach} \\
        \B{impo} & : & seq \B{Impo} \\

@@

        \B{op} & : & seq \B{Oper} 

@@

        \B{mach} & : & \B{Mach} \\
        \B{pre} & : & opt \B{Name} 
        \multicolumn{3}{|c|}{Machine: \B{Mach}} \\

@@

        \B{impl} & : & \B{Impl} \\

@@

        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\

@@

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def} 
      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{ne} \alt \llvm{sgt} \alt \llvm{sge} \alt \llvm{slt} \alt \llvm{sle} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}
  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
    and \nt{name} correspond respectively to integer types, integer literals,
    labels and names. Choices are separated by \alt and optionally delimited by
    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
    elements of the annotated entity.}
  \label{fig:llvm-grammar}
\end{figure}

@@

To conclude, we make no assumption on the existence of a library to obtain
resources managed by the operating system, such as dynamic memory allocation. In
the absence of such resource, all data must be allocated either statically, or
on the current stack frame, by using the \llvm{alloca} instruction.

@@

  The parameter list of the functions thus produced contains one item for each
  input and output of the corresponding operation, and one item of type
  \llvm{\%M\$state\$*}, to pass the address of the structure encoding the
  state of the implementation and its modification by the callee. The return
  type is always \llvm{void}.
\end{enumerate}
The translation is designed to support modular B development. In a B
implementation, one may reuse existing components, by instantiating (the B term
is \emph{importing} so-called library machines. Such library machines have their
own development and implementation. Borrowing terminology from the
object-oriented world, one may view library machines as classes, and the B
implementation as an object that contains instances of those classes. In
practice, different from most OO languages, such instances are not references,
but the actual objects. We consider this necessary as dynamic memory allocation
is precluded.

To translate the instantiations of library machines in an implementation, the
translator needs to have access to the following information on those machines
and their respective implementation:
\begin{itemize}
\item for each library machine \B{M}, the corresponding implementation, denoted
  \B{M.impl}.
\item for such implementation, the type of the state and the signature of the
  operations.
\end{itemize}
Note that, when translating the implementation of a library machine, no instance
shall be generated. It is in the implementation of the main design, that we call
the \emph{top-level} component, that such instances are created.

Figure~\ref{fig:skel} summarizes the translation principles by providing the
structure of the LLVM implementation thus produced.

\begin{figure}
  \begin{center}
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}} \\
      \textit{if this is the top-level component, generate} \\
      \> \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer}
    \end{tabbing}
  \end{center}
  \caption{Structure of the LLVM module produced from a B implementation \B{M}.}
  \label{fig:skel}
\end{figure}

@@

\paragraph{Notations.} Code generation is defined with a set of rules defined by
structural recursion. For each grammatical construct, a code generation function
is defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
instructions, $\trad{~}{\B{Expr}}$ for expressions and so forth. Each such
function usually produces a text corresponding to bits of LLVM source code. It
may have additional parameter and results. The symbol $\sep$ is used to clearly
separate composed results. Auxiliary functions are introduced for
conveniency. To abbreviate the presentation, we denote
$\trad{~}{\ListOf{\B{A}}}$ as the translation function for sequences of \B{A}
elements, defined as the concatenation of the application of $\trad{~}{\B{A}}$
to the elements of the sequence.

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name.  Also, we define
three auxiliary functions responsible to produce such identifiers of,
respectively, the type representing the state of a B module \B{n}, its
initialization function, and the object carrying its state:
\begin{align*}
\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
\end{align*}
We assume an unlimited supply of labels and local variables: we use $\lbl$ to
return a fresh LLVM label and $\variable$ for a fresh local variable name.

@@

The translation of a B implementation to LLVM, following the template sketched
in Figure~\ref{fig:skel}, is specified by the following rule, where \B{n} is the
root of the implementation representation and $\top^?$ a boolean indicating if
\B{n} is the implementation of a top-level component or of a library machine:
\begin{align*}
  \lefteqn{\trad{\B{n}, $\top^?$}{\B{Impl}} \isdef } \\
  & \llvm{"\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{State}}}_{def}}} & \text{defines the type of imported modules} \\
  & \llvm{\state(\B{n}) = type \PH{\trad{n}{State}_{expr}}} & \text{defines the type of current module} \\
  & \llvm{\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{Bhv}}}_{decl}}} & \text{declares of imported modules operations} \\
  & \llvm{\PH{\trad{\B{n.init}}{\B{Init}}}} & \text{defines the initialization function} \\
  & \llvm{\PH{\trad{\B{n.op}}{\ListOf{\B{Oper}}}}} & \text{defines functions implementing operations}\\
  & \llvm{\PH{\trad{\B{n}, $\top^?$}{\B{Inst}}}"} & \text{creates instance (conditional)}
\end{align*}

To handle the importation of a machine, one needs to produce the definition of
the type for the state of its implementation. Since this implementation may reference
other modules, the definition is recursive (parameter $A$ is an accumulator to avoid
duplicating such definitions):
\begin{align*}
  \trad{\B{n}, $A$}{\B{State}}_{def} \isdef 
  & \IF \B{n} \not\in A \THEN \\
  & \quad \LET \B{m} = \B{n.mach} \IN \\
  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
  & \quad \quad \quad \llvm{"\PH{\trad{\B{i.impo}, $A \cup \{ \B{n} \}$}{\ListOf{\B{State}}}_{def}}} \\
  & \quad \quad \quad \llvm{\PH{\state(\B{m})} = type \PH{\trad{i}{State}_{expr}} \nl"}
\end{align*}

The following rules specify how the type representing the state of an
implementation is generated. It is a structure type, with one element per
imported machine, and one element per variable. The definition of such type is
just the enumeration of the type of its elements (the individual elements are
referenced, not by name, but by position inside the structure):
\begin{align*}
  \trad{\B{n}}{\B{State}}_{expr} \isdef & \llvm{"\{ \PH{\trad{\B{n.impo}}{\ListOf{\B{Impo}}}_{expr}},  \PH{\trad{\B{n.var}}{\ListOf{\B{Vari}}}_{expr}} \}"} \\
  \trad{\B{n}}{\B{Impo}}_{expr} \isdef & \state(\B{n.mach}) \\
  \trad{\B{n}}{\B{Vari}}_{expr} \isdef & \trad{\B{n.type}}{\B{Type}}
\end{align*}

The code generation for the functions implementing the initialization and
operations of imported machine is specified by the following rules. Not that
this definition needs not be recursive, as only the operations of a directly
imported machine may be called in an implementation. Again, parameter $A$ plays
the role of an accumulator to avoid (harmless, but useless) duplicate
declarations.
\begin{align*}
  \trad{\B{n}, $A$}{\B{Bhv}}_{decl} \isdef 
  & \IF \B{n} \not\in A \THEN \\
  & \quad \LET \B{m} = \B{n.mach} \IN \\
  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
  & \quad \quad \quad \llvm{declare void \PH{\init(\B{m})}(\PH{\state(\B{m})} *) \nl} \\
  & \quad \quad \quad \llvm{\PH{\trad{\B{m.op}}{\ListOf{\B{Oper}}}_{decl}}"} \\
  \trad{\B{n}}{\B{Oper}}_{decl} \isdef &
  \llvm{"declare void \PH{\Global(\B{n})}(\PH{\state(\B{n.root})}, \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}}, \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}})"}\\
  \trad{\B{n}}{\B{inp}}_{decl} \isdef & \trad{\B{n.type}}{\B{Type}} \\
  \trad{\B{n}}{\B{out}}_{decl} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
\end{align*}

The last rule is responsible for producing the code creating the instance when the
implementation is that of the top-level component of a design:
\begin{align*}
\lefteqn{  \trad{\B{n}, $\top^?$}{\B{Inst}} \isdef } \\
  & \IF \top^? \THEN \llvm{"\PH{\self(\B{n})} = common global \PH{\state(\B{n})} zeroinitializer \nl"} \\
  & \ELSE \llvm{""}
\end{align*}

\section{Translation: Data}
\label{sec:data}
